---
title: "R Homework"
author: "prita, heva, supria"
date: "March 29, 2018"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Visualization in R : Long-term monitoring of a rodent community

Our homework is analyzing database of a long term monitoring for rodent community in Chihuahuan Desert ecosystem near Portal, Arizona, from 1977 to 2000. At this site, 24 experimental plots were established in 1977 and divided among controls and experimental manipulations. 

The long-term data for the rodent community at the Portal Project has been used to address a variety of questions including
<u>
  <li>Tool to monitor the community- and population-level dynamics of desert rodents and competitive interactions among rodent species.</li>
  <li>Responses of rodents to climatic variability.</li>
  <li>The long-term stability and dynamics of a desert rodent community.</li>
</u>  


This is an R Homework documentation. The process of data visualisation using R can be divided into four steps : 
<ul>
  <li>Load the library</li>
  <li>Read the data file</li>
  <li>Connect to database sqlite</li>
  <li>Data Cleansing</li>
  <li>Data Plotting</li>
</ul>

<strong>Load the library</strong>

```{r load, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(gridExtra)
library(ggplot2)
library(dbplyr)
library(RSQLite)
library(DBI)
```


<strong>Reading the data file</strong>

```{r read}
surveys_combined <- read.csv("data/combined.csv")
```

<strong>Connect to database sqlite</strong>

```{r connect}
mammal_db <- dbConnect(dbDriver("SQLite"), "data/portal_mammals.sqlite")
```

<strong>Data Cleansing</strong>

```{r clear1}
surveys_combined_clear<- surveys_combined %>% filter(!is.na(sex), 
                                                         sex != "",
                                                         !is.na(hindfoot_length), 
                                                         hindfoot_length != "",
                                                         !is.na(weight), 
                                                         weight != "") 
```

```{r clear2}
surveys_db <- dbGetQuery(mammal_db, "select surveys.*, species.species, species.taxa, species.genus,
                                      plots.plot_type
                                      from surveys, species, plots
                                      where surveys.species_id = species.species_id
                                      and surveys.plot_id = plots.plot_id
                                      and surveys.sex <> 'NA'
                                      and surveys.hindfoot_length <> 'NA'
                                      and surveys.weight <> 'NA'
                                    ")
```

<strong>Data Plotting</strong>

```{r plot1}
#create line chart plot type per year
year_plot_type <- surveys_combined_clear %>% group_by(year, plot_type) %>% tally()
line_chart <- ggplot(year_plot_type, aes(x=year, y=n, color=plot_type)) + 
  geom_line() + xlab("Year") + ylab("Plot Type")

#create bar chart sex per year
year_sex <- surveys_combined_clear %>% group_by(year, sex) %>% tally()
bar_chart <- ggplot(year_sex, aes(x=year, y=n, color=sex)) + 
  geom_bar(stat="identity") + xlab("Year") + ylab("Sex")

#put chart to grid
timeseries_plot <- grid.arrange(bar_chart, line_chart, ncol=2, widths=c(4,6))

#save plot into image
ggsave("image/plot1.jpg", timeseries_plot, width=10, dpi=300)
```


```{r plot2}
#create boxplot chart weight per plot_type
boxplot_chart_weight <- ggplot(surveys_combined_clear, aes(x=plot_type, y=weight))+
  geom_boxplot()+xlab("plot type")+ylab("Weight (g)") +  geom_jitter() +
  scale_y_log10()

#create boxplot chart hindfoot length per plot_type
boxplot_chart_length <- ggplot(surveys_combined_clear, aes(x=plot_type, y=hindfoot_length))+
  geom_boxplot()+xlab("plot type")+ylab("Length") + geom_jitter()
  scale_y_log10()

#put chart to grid
frequency_plot <- grid.arrange(boxplot_chart_weight, boxplot_chart_length, ncol=2, widths=c(5,5))

#save plot into image
ggsave("image/plot2.jpg", frequency_plot, width=10, dpi=300)
```



## Analyzing the Plot

<strong>To Do List</strong>
<ul>
<li>find the correlation and p-value & R-square</li>
<li>tidy up the RMarkdown documentation</li>
</ul>

To be continued ... 
